Live Functional Programming with Typed Holes

### 2018-04-04 intro.tex
intro nits; up to and including 'prior approaches'

In a traditional programming environment, the user iteratively edits the program
text, compiles and runs the program, and views the output.
%
To make this workflow more usable and efficient, \emph{live programming
environments} ``promise to narrow the temporal and perceptive gap between
program development and code execution''~\cite{burckhardt2013s} with
fine-grained interleaving of editing and evaluation to varying
degrees~\cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}.

@@

Despite these advances, however, there is a fundamental limitation which hinders
the full realization of live programming: traditional programming languages
cannot evaluate \emph{incomplete programs}, \ie{}, programs that are either
syntactically ill-formed or ill-typed.
%
Without a well-defined semantics to dictate the treatment of incomplete
programs, various editor services may be provided in ad-hoc, restricted
forms---if available, at all---during the periods of time between complete
program states.
%
%% ``flickering in and out''
%
%% In other words, the ``Edit-Compile-Run'' cycle routinely fails (either in the
%% first or second steps) during the program editing process.
%
The resulting editor experience includes services ``flickering in and out,''
thus introducing ``temporal and perceptive gaps'' because programmers often
leave a program malformed or ill-typed for extended periods of time, \eg{}~as
they think about what to enter at the cursor, or as they work on a different
part of the program.
%
These times are, arguably, when rich, live editor services would be most helpful
to the user.

@@

\paragraph{Exceptions as Expression Holes}
%
One common approach for emulating expression holes is to use exceptions (such as
\verb+Debug.crash "not yet implemented"+ or \verb+raise NotYetImplementedException+).
%
Some languages even provide syntax for holes
%
(written \verb+_+ in GHC Haskell~\cite{XXX},
\verb+?+ in Agda~\cite{XXX}, and
\verb+{x}+ in Idris~\cite{XXX})
%
which, semantically, behave similarly to exceptions.
%
A placeholder exception serves to appease the static type checker---because an
error expression is allowed to have any type required by the surrounding
context---allowing the program to run, while preserving dynamic type safety by
simply halting execution as soon as the expression appears in evaluation
position.

Although useful, this ad-hoc approach has limitations.
%
Because the program must halt as soon as a placeholder expression is reached (in
order to ensure type safety), the programmer cannot benefit from seeing how
other parts of the program, which do not depend on the missing expression,
continue to evaluate.
%
Another limitation is that exceptions typically carry nothing more than the
stack trace that led to the error.
%
But there are other forms of information that could be useful as the programmer
works to fill the hole---\eg{}, the dynamic environment around the failed
expression, and the subsequent computations that depend on the missing value
(which could only be deduced if evaluation continued).
%
Thus, the dynamic semantics of exceptions do not adequately address the
particular needs for expression holes.

@@

>>>>

In a traditional programming environment, the user iteratively edits the program
text, compiles and runs the program, and views the output.
%
To make this workflow more usable and efficient, \emph{live programming
environments} ``promise to narrow the temporal and perceptive gap between
program development and code execution''~\cite{burckhardt2013s} with
fine-grained interleaving of editing and evaluation~\cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}.

@@

Despite these advances, however, there is a fundamental limitation which hinders
the full realization of live programming: traditional programming languages
cannot evaluate \emph{incomplete programs}, \ie{}, programs that are either
syntactically ill-formed, or ill-typed.
%
Without a well-defined semantics to dictate the treatment of incomplete
programs, various editor services may be provided in ad-hoc, restricted
forms---if available at all---during the periods of time between complete
program states.
%
%% ``flickering in and out''
%
%% In other words, the ``Edit-Compile-Run'' cycle routinely fails (either in the
%% first or second steps) during the program editing process.
%
The resulting editor experience includes services ``flickering in and out,''
thus introducing ``temporal and perceptive gaps'' because programmers often
leave a program malformed or ill-typed for extended periods of time, \eg{}~as
they think about what to enter at the cursor, or as they work on a different
part of the program.
%
These times are, arguably, when rich, live editor services would be most helpful
to the user.

@@

\paragraph{Exceptions as Expression Holes}
%
In lieu of proper language and tool support for expression holes,
today's programmers commonly emulate them with placeholder exceptions, such as
\verb+Debug.crash "not yet implemented"+ or
\verb+raise NotYetImplementedException+.
%
Some languages even provide syntax for these kind of ``holes''
%
(written \verb+_+ in GHC Haskell~\cite{XXX},
\verb+?+ in Agda~\cite{XXX}, and
\verb+{x}+ in Idris~\cite{XXX})
%
which, semantically, behave similarly to exceptions.

Such exceptions serve to appease the static type checker---because an
error expression is allowed to have any type required by the surrounding
context.
%
Thus, this mechanism permits the program to run, while preserving dynamic type safety, and
simply halting execution as soon as the expression appears in evaluation
position.

Although useful, an exception-based mechanism for ``program hole'' has limitations.
%
First,
the program must halt as soon as this expression is reached (ensuring type safety), and consequently, the programmer cannot benefit from seeing how
\emph{other parts} of the program, which may not depend on the missing expression,
continue to evaluate.

Second, exceptions typically carry nothing more than the
stack trace that led to the error.
%
But there are other forms of information that could be useful as the programmer
works to fill the hole---\eg{}, the dynamic environment around the failed
expression, and the subsequent computations that depend on the missing value
(which could only be deduced if evaluation continued).
%
Thus, the dynamic semantics of exceptions do not adequately address the
particular needs for expression holes.

@@

%%%%%%%%%
%% Matt says: I added the remarks below; My intention is to tie the
%% last bit into the ``edit and resume'' feature, without making a
%% huge deal about it.

Finally, exceptions are often thrown with the intent that some calling
context will possibly ``handle'' them, and will thus resume ordinary
(non-exception-based) program control flow.
%
Meanwhile, it never really makes sense to handle a ``not yet
implemented'' exception, save for \emph{editing the program to further
  complete it}, which is beyond the scope and control of ordinary
exception handlers, and (ordinary) programs' execution behavior.
%
Indeed, this is one (of many) ``perceptual gaps'' for live programming
environments to close.
%
In this work, we refer to the programmer-user interaction pattern that
follows such exceptions as ``edit and resume''.
